Cross-relaxation in dense ensembles of NV centers and application to magnetometry \\ \medskip Relaxation croisées d'ensembles de centres NV et application à la magnétométrie

### 2022-10-20 Manuscrit Thèse/These.tex
tata

In this introductory chapter, we will first cover the material aspects of NV centers by focusing on the creation process of synthetic diamonds and the required steps to create NV defects inside them. We will then cover the fundamentals of NV physics, including the NV center optical and spin properties, basic experiments with NV centers and the various relaxations times ($T_1$,$T_2$,$T_2^*$) which characterizes the NV center's spin. Finally we will present the dipole-dipole interaction between two spins and the concept of cross-relaxation.

\section{The NV center as a physical object}
\begin{figure}[h!]
\centering
\includegraphics[width=0.9\textwidth]{chapter 1/Figures/diamant_et_NV}
\caption{From the diamond to the NV center. a) Drawing of a bulk diamond\footnote{In this manuscript, we will consider any samples above $\sim$ 100 $\mu$m to be bulk.}. b) Drawing representing a fluorescence microscopy image of a diamond surface. The red fluorescent spots symbolize NV centers photoluminescence. c) Zoom on the crystalline structure of an NV center. Black balls represent carbon atoms}
\label{diamond+NV}
\end{figure}
Before studying the NV centers properties, we must first discuss the diamond itself and answer these questions: How are diamond created, what kind of defects are there in diamond, and how to specifically create NV defects. The field of material science related to diamonds and NV centers is a rich and complex one, and this section will only touch upon basic notions regarding these subjects.

@@

The rates $\gamma$ of the various decay processes, represented in Fig. \ref{ISC} by the time constant $\tau=1/\gamma$, strongly depend on the spin projection $m_s$. Importantly, the likeliness to undergo a non-radiative decay is $\sim 4$ times\footnote{$\frac{15+91}{15+11}\approx 4$} more likely when starting from the $\ket{\pm 1}$ states than when starting from the $\ket{0}$ state. Similarly, the singlet state is $\sim 2.4$ times\footnote{$\frac{492}{204} \approx 2.4$} more likely to decay in the $\ket{0}$ state than in the $\ket{\pm 1}$ states.

These spin selective intersystem crossings mean that an optical pumping scheme in the $\ket{m_s=0}$ spin state takes place under constant green excitation. At optical saturation, the polarization of the $\ket{0}$ state can reach $70-90\ \%$ \citep{gupta2016efficient}, which is equivalent to a thermal polarization (for B=0) with a spin temperature of $\approx 65\ \rm mK$ \footnote{\begin{equation*}
T=\frac{h D}{k_B \ln (\frac{1-P}{2P})} \approx 65\ \rm mK,
\end{equation*}
where $D$ is th ZFS factor described in eq. (\ref{eq. spin elec}) and $P\approx 0.8$ is the spin polarization defined as the population in the $\ket{0}$ state.} while the diamond itself can remain at room temperature.

@@

\section{Basic operations with NV centers}

This section introduces two elementary operations with NV centers that are Rabi oscillations and optically detected magnetic resonance (ODMR).

Rabi oscillations is the basis for any kind of coherent qubit manipulation while ODMR is used to find the transition frequencies of the NV centers, as well as being a basic magnetometry protocol.

This section also covers the experimental setup used for these two operations and for many other NV experiments, as well as the concept of longitudinal and transverse magnetic field with regard to the NV principal axis.

\subsection{Experimental setup}
\label{Sec setup cahp 1}
\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{chapter 1/Figures/shema_exp}
\caption{Confocal microscopy setup detailed in main text}
\label{setup ch1}
\end{figure}
The three most common experimental degrees of freedom to consider for NV experiments are: the optical illumination, the microwave field and the external magnetic field. Other parameters such as the crystal temperature, the electric field or the crystal stress can also be controlled for specific applications.

Finally the magnetic field is generated either with a permanent magnet (up to $\sim$ 200 mT) or with a homemade electromagnet (up to $\sim$ 30 mT) alimented via a low freqency amplifier (Leybold power function generator 522663). Both the permanent magnet and the electromagnet can be  attached to a rotation stage for precise alignment.
\subsection{Rabi Oscillations}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/Rabi}
\caption{Rabi oscillations. (a) pulse sequence used to measure Rabi oscillations. (b) Experimental data on sample CVD-pink (see sample details in appendix \ref{Appendix samples}).}
\label{Rabi}
\end{figure}

The general technique for experiments involving NV spins proceeds as follow: 
\begin{itemize}
\item First the spins are polarized. This is done with a green laser pulse ($1-100$ $\mu$s depending on the laser intensity) which polarizes the spins in the $\ket{m_s=0}$ state at $\sim$ 80 \%.
\item  Then the spins are manipulated. This process often involves a microwave field resonant with one of the spin transitions.
\item Finally the spins are read out. The simplest spin readout for NV centers consists in looking at the PL right after the laser is turned on: if the spin was in the $\ket{m_s=0}$ state, the initial PL will be high, whereas if the spin was in a $\ket{m_s=\pm 1}$ state, the initial PL will be low. For longer exposure time, the spins end up re-polarized in the $\ket{m_s=0}$ state and no information can be gained on the initial spin state anymore. The optimal readout time is of the same order as the spin polarization time ($1-100$ $\mu$s).
\end{itemize}

Fig. \ref{Rabi}-a) shows the pulse sequence used to observe Rabi oscillations with NV centers. In this case, the spins manipulation simply consists in a microwave pulse of variable duration $\tau$. The microwave frequency or the magnetic field needs to be adjusted so that at least one NV spin transition (either $\ket{0} \to \ket{+1}$ or $\ket{0} \to \ket{-1}$) is resonant with the microwave field.

Fig. \ref{Rabi}-b) shows the result of such an experiment on a spin ensemble. The PL signal oscillations correspond to the oscillations of the spin population between the bright $\ket{0}$ state and the dark $\ket{-1}$ state. For long microwave pulses ($>$ 1 $\mu$s typically), the spins reach a steady state with  $\sim$ 50 \% population in both the $\ket{0}$ and $\ket{-1}$ states.

\subsection{Optically detected magnetic resonance}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/ODMR_basique}
\caption{ODMR protocol. (a) Pulse sequence used for CW ODMR: the microwave is kept on during the whole sequence while the frequency is scanned from $f_{\rm start}$ to $f_{\rm end}$. (b) ODMR spectrum with no magnetic field on sample ADM-15-4.}
\label{ODMR basique}
\end{figure}

As we saw in the last section, the steady state of the NV center spin with a resonant microwave field is a statistical mixture of 50 \% $\ket{0}$ and 50 \% $\ket{\pm 1}$ \footnote{We are assuming here that the Rabi frequency is greater than the laser polarizing rate, which is often the case in practice}. If the microwave field is not resonant with a spin transition however, the steady state is predominantly $\ket{0}$ due to the spin polarization caused by the laser. The NV PL will therefore be lowered when the scanning microwave field becomes resonant with one of the NV spin transitions.

Fig. \ref{ODMR basique}-b) shows an example of ODMR spectrum on an NV ensemble with no external magnetic field. The reason for the two bumps instead of a single one centered on $D=2870\ \rm MHz$ is the splitting of the $\ket{+1}$ and $\ket{-1}$ levels occasioned by the strain and local electric field, which will be further discussed in chapter 4.

\subsubsection{ODMR with lock-in detection}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/ODMR_lockin}
\caption{ODMR protocol with lock-in detection. (a) Simplified confocal microscope setup with a lock-in amplifier (LIA) modulating the mircowave and demodulating the PL. (b) Demodulated ODMR spectrum with no magnetic field on sample ADM-15-4.}
\label{ODMR lockin}
\end{figure}

Because the ODMR protocol delivers a continuous signal (the PL) as a function of a periodic excitation (the microwave field), it is well suited to amplitude or frequency modulation and lock-in detection.

The principle of lock-in detection will not be covered here, the key idea is that lock-in detection allows to translate the signal spectrally in a region less prone to noise. In this case, the ODMR signal is a DC measurement and is therefore prone to slow freqency noises such as thermal or mechanical drifts and laser amplitude fluctuations. The microwave field can thus be modulated (in amplitude, frequency or phase) at a frequency fast enough to mitigate the low freqency noises, but slow enough compared to the NV photodynamics.

\subsection{ODMR with NV ensemble: longitudinal and transverse magnetic field}
\label{sec champs transverse}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/ODMR_8_classes}
\caption{(a) Geometry of the NV center and definition of $B_z$ and $B_\perp$. (b) Representation of the 4 possibles orientations (``classes") of the N$-$V axis, corresponding to the 4 $\langle 111 \rangle$ directions of the diamond cubic cell. (c) Simulation of the evolution of the transition frequencies for each class of NV centers as a function of the amplitude of the applied magnetic field ($\textbf{B}$ orientation chosen arbitrarily). (d) ODMR spectrum on sample ADM-150-1 for a magnetic field of $\sim 90\ \rm G$.}
\label{ODMR 8 classes}
\end{figure}

We mentioned previously that the NV spin Hamiltonian (eq. (\ref{NV spin Hamiltonian basic})) was anisotropic due to the fine structure term $D S_z^2$ where $z$ is the direction of the N$-$V axis in the crystal. The magnetic field is thus often decomposed between the longitudinal magnetic field $B_z$ and the transverse magnetic field $B_\perp=\sqrt{B_x^2+B_y^2}$ (see Fig. \ref{ODMR 8 classes}-a).

For weak magnetic fields ($\gamma |\textbf{B}| \ll D$), the energy of each of the three spin Hamiltonian eigenstates $\ket{0}$, $\ket{+1}$ and $\ket{-1}$\footnote{The notation of $\ket{0}$, $\ket{\pm 1}$ to designate the spin Hamiltonian eigenstates is slightly abusive as $S_z$ does not commute with $\mathcal{H}$ when $B_\perp \neq 0$. The true eigenstates however remain close to $\ket{0}$, $\ket{\pm 1}$ as long as $\gamma_e B_z > \frac{(\gamma_e B_\perp)^2}{D}$.} can be approximated in a perturtbative approach to: %cf AN.py
\begin{align}
\nu_0&= -\frac{(\gamma_e B_\perp)^2}{D} \\
\nu_{-1} &= D - \gamma_e B_z + \frac{(\gamma_e B_\perp)^2}{2D} \\
\nu_{+1} &= D + \gamma_e B_z + \frac{(\gamma_e B_\perp)^2}{2D}
\end{align}

Fig. \ref{ODMR 8 classes}-b) represent the 4 possible orientation of the N$-$V axis in a diamond single crystal. We refer to these orientations as ``classes" since NV centers belonging to the same class will perceive the same transverse and longitudinal magnetic field, and thus will have the same spin Hamiltonian. For large NV ensembles (a confocal spot can contain more than $10^9$ NV centers in dense ensemble), we can consider that each class is equally represented \footnote{It is possible to create NV centers with a preferential orientation \citep{lesik2014perfect} but this requires specific conditions during the growth process of the diamond} and that each class is responsible for two lines in the ODMR spectrum (one for the $\ket{0} \to \ket{-1}$ transition and one for the $\ket{0} \to \ket{+1}$ transition).

Fig. \ref{ODMR 8 classes}-c) shows a simulation of the evolution of the transition frequencies $\nu_{+1} - \nu_{0}$ and $\nu_{-1} - \nu_{0}$ for all 4 classes as a function of the magnetic field amplitude, with the magnetic field orientation fixed in an arbitrary direction. Fig. \ref{ODMR 8 classes}-d) shows an ODMR spectrum with NV ensembles. 8 lines are visible corresponding to the two transitions of the 4 classes. Based on the position of these 8 lines, it is possible to determine the projection of the magnetic field on each of the 4 N$-$V axes, and thus to reconstruct the full 3D magnetic field with respect to the diamond crystalline axes.

The relaxation times are generally decomposed between the longitudinal relaxation time $T_1$ which corresponds to the characteristic time of the return to thermal equilibrium for the populations (diagonal elements of the density matrix), and the transverse relaxation time $T_2$\footnote{We follow here the NMR nomenclature} which corresponds to the relaxation of the spin state coherences (non diagonal elements of the density matrix).

The transverse relaxation time, also called coherence time or dephasing time, is often further decomposed between the inhomogeneous dephasing time $T_2^*$ and the homogeneous dephasing time $T_2$. The $T_2^*$ time includes inhomogeneous dephasing processes such as the inhomogeneity in space (when multiple qubits are probed) and time of the external fields. The $T_2$ time corresponds to the coherence time where some of these inhomogeneities have been corrected, for instance with a Hanh echo sequence \citep{hahn1950spin} or a dynamical decoupling scheme \citep{naydenov2011dynamical}. While $T_2^*$ is an intrinsic property of the sample studied, $T_2$ depends on the rephasing scheme used. More careful nomenclatures will label the $T_2$ times as $T_{2,\rm echo}$ or $T_{2,\rm DD}$ \citep{de2021materials} for instance. The relaxation times are ordered as follows:
\begin{equation}
T_2^* < T_2 < 2 T_1.
\end{equation}

The relaxation times of NV centers depend mainly on their environment, and can vary by several order of magnitude depending on the sample used. The following paragraphs will cover the measurement processes of the different relaxation times and the typical values for the samples used in this manuscript.

\subsection{NV spin $T_1$}
\label{sec T1 NV}
\subsubsection{Measurement protocol}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/T1_basique}
\caption{Basic $T_1$ measurement scheme. (a) Pulse seqeunce used. (b) Experimental data obtained on sample CVD-pink for low laser intensity. The orange line corresponds to an exponential fit with time constant $\tau=4.0\ \rm ms$.}
\label{T1 basique}
\end{figure}

$T_1$ is the simplest relaxation time to measure because the spin populations are directly correlated to the NV centers PL. Fig. \ref{T1 basique} shows the pulse sequence used and an experimental result of the $T_1$ sequence. The spin lifetime $T_1$ is one of the only spin property that can be measured without a resonant microwave.

The sequence shown in Fig.\ref{T1 basique} can be improved by implementing a common mode rejection technique described in Fig. \ref{T1 soustraction}-a). This protocol works as follow: first, a basic $T_1$ sequence is recorded, which we will denote $S_1$. Then the sequence is repeated with an additional $\pi$-pulse on one the NV transitions right before the readout. This $\pi$-pulse will project the remaining $\ket{0}$ population difference into a $\ket{\pm 1}$ state. The second readout is labeled $S_2$, and the final spin state readout is $S_1 -S_2$ which rejects every contribution to the PL outside of the spin relaxation.

Fig. \ref{T1 soustraction}-b) shows the experimental result of the $S_1$ and $S_2$ sequences on the same sample that was used in Fig. \ref{T1 basique}. The difference between the $S_1$ data in Fig. \ref{T1 soustraction}-b) and the the data in Fig. \ref{T1 basique}-b) comes from the laser intensity used. In Fig. \ref{T1 soustraction}, the laser intensity was high enough ($> 1\ \rm{mW}$) that photoionization between NV$^-$ and NV$^0$ becomes significant. Furthermore, for dense ensemble, a charge recombination mechanism operates even in the dark \citep{giri2018coupled, giri2019selective} (most likely due to charge tunneling between impurities \citep{choi2017depolarization}) . On the other hand, the laser intensity used in Fig. \ref{T1 basique} was small enough ($< 10\ \mu \rm{W}$) that the charge state state photodynamics contribution was negligible compared to the spin depolarization.

Fig. \ref{T1 soustraction}-c) Shows the subtraction of the $S_1$ and $S_2$ signal. Despite the complex shape of the initial $S_1$ and $S_2$ signals, $S_1-S_2$ is nicely fitted with an exponential decay similar to the results found in Fig. \ref{T1 basique}.

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/T1_protocole_soustraction}
\caption{Common mode rejection $T_1$ measurement scheme. (a) Pulse seqeunce used. (b) PL signals $S_1$ and $S_2$ obtained on sample CVD-pink with a high laser intensity. (c) Subtraction $S_1-S_2$ versus dark time $\tau$. The experimental data is fitted with an exponential decay of time constant $\tau=4.1\ \rm ms$.}
\label{T1 soustraction}
\end{figure}

We distinguish three different origin for the the $T_1$ relaxation process:
\begin{itemize}
\item \textbf{Spin-phonon relaxation:} the spin states are weakly coupled to the crystal phonon bath through spin-spin and spin-orbit interactions \citep{norambuena2018spin}, which can result in a phonon-induced spin relaxation. The dominant relaxation mechanism depends on the  temperature regime, with the two-photon Raman process believed to be dominant at room temperature \citep{takahashi2008quenching, jarmola2012temperature} while the dominant process as at cryogenic temperature ($4\sim100$ K) is believed to be a two phonons Orbach process \citep{redman1991spin, norambuena2018spin}. The spin-phonon relaxation time strongly depends on the crystal lattice temperature, with relaxtion times of the order $T_1 \sim 5\ \rm ms$ at room temperature, and up to 8 hours at for mili-K temperatures \citep{astner2018solid}.

\item \textbf{Electric and magnetic field noises:} both electric and magnetic fields can directly excite the $\ket{0} \to \ket{\pm 1}$ spin transition \citep{udvarhelyi2018spin}. As a result, magnetic and electric field noise can reduce the spin $T_1$ if their spectrum overlap with the NV transitions. This is particularly important for near surface spins because of the strong noise caused by electronic defects on the surface \citep{sangtawesin2019origins}. The $T_1$ of near surface NV center can be as low as $\sim 10\ \mu$s, but the surface effects are only effective at very short range (within $\sim 10\ \rm nm$ of the surface).

\item \textbf{Resonant dipolar interaction with other paramagnetic impurities:} Dipolar interaction betwen two spins will be detailed more thoroughly in the following section. Dipole-dipole interaction between NV centers and surrounding spins is responsible for both transverse and longitudinal spin relaxation, with the added condition of energy matching (resonance) for longitudinal relaxation. The $T_1$ time associated with dipole-dipole interaction strongly depends on the spin concentration and can be as low as $\sim 25 \mu$s \citep{hall2016detection}.
\end{itemize}

The $T_2^*$ time is the characteristic time for the loss of quantum coherence of a freely evolving system. It is generally defined as the decay time of of a Ramsey interferometry measurement \citep{barry2020sensitivity}.

Fig. \ref{Ramsey} shows the Ramsey pulse sequence and an  example obtained on a spin ensemble. The microwave frequency is detuned by $\sim 300\ \rm kHz$ compared to the spin central Larmor frequency which causes oscillations in the signal. $T_2^*$ is the decay time of the envelope of the Ramsey fringes. Unfortunately, it is hard to correctly fit the envelope of a signal with so few fringes. Increasing the number of fringes with our current microwave and optical setup might prove hard, as further microwave detuning would decrease the signal contrast, and would be ultimately limited by the microwave pulses resolution.

An alternative measurement of $T_2^*$ can be performed by looking at the linewidth of an ODMR spectrum. Indeed, ideally the ODMR spectrum should be the Fourier transform of the Ramsey signal (also called free induction decay in NMR literature). In practice however, both methods suffer from experimental noises and imperfections, with Ramsey-type methods being more sensitive to microwave inhomogeneities \citep{barry2020sensitivity} while ODMR is prone to laser and microwave power broadening \citep{dreau2011avoiding}.

The method chosen to extract $T_2^*$ values in this manuscript was ODMR with low microwave power ($\Omega_{\rm Rabi} \ll 1/T_2^*$) and low laser intensity (compared to the the NV optical saturation).

@@

>>>>

In this introductory chapter, we will first cover the material aspects of NV centers by focusing on the creation process of synthetic diamonds and the required steps to create NV defects inside them. We will then cover the optical and spin properties of single NV centers, and introduce the notions of dipole-dipole interaction and cross-relaxation. We will then cover basic experiments with NV centers and introduce the experimental tools required to characterize the NV center's spin.

\section{The NV center as a physical object}
\begin{figure}[h!]
\centering
\includegraphics[width=0.9\textwidth]{chapter 1/Figures/diamant_et_NV}
\caption{From the diamond to the NV center. a) Drawing of a ``bulk" (macroscopic) diamond. b) Drawing representing a fluorescence microscopy image of a diamond surface. The red fluorescent spots symbolize NV centers photoluminescence. c) Zoom on the crystalline structure of an NV center. Black balls represent carbon atoms}
\label{diamond+NV}
\end{figure}
Before studying the NV centers properties, we must first discuss the diamond itself and answer these questions: How are diamond created, what kind of defects are there in diamond, and how to specifically create NV defects. The field of material science related to diamonds and NV centers is a rich and complex one, and this section will only touch upon basic notions regarding these subjects.

@@

The rates $\gamma$ of the various decay processes, represented in Fig. \ref{ISC} by the time constant $\tau=1/\gamma$, strongly depend on the spin projection $m_s$. Importantly, the likeliness to undergo a non-radiative decay is $\sim 4$ times\footnote{$\frac{15+91}{15+11}\approx 4$} more likely when starting from the $\ket{\pm 1}$ states than when starting from the $\ket{0}$ state. Similarly, the singlet state is $\sim 2.4$ times\footnote{$\frac{492}{204} \approx 2.4$} more likely to decay to the $\ket{0}$ state than to the $\ket{\pm 1}$ states.

These spin selective intersystem crossings mean that an optical pumping scheme in the $\ket{m_s=0}$ spin state takes place under constant green excitation. At optical saturation, the polarization of the $\ket{0}$ state can reach $70-90\ \%$ \citep{gupta2016efficient}. This is equivalent to a spin-temperature $\approx 65\ \rm mK$ \footnote{\begin{equation*}
T=\frac{h D}{k_B \ln (\frac{1-P}{2P})} \approx 65\ \rm mK,
\end{equation*}
where $D$ is th ZFS factor described in eq. (\ref{eq. spin elec}) and $P\approx 0.8$ is the spin polarization defined as the population in the $\ket{0}$ state.}.

@@

@@

\section{Optically detected magnetic resonance (ODMR)}

This section introduces one of the building block of almost any NV experiment which is the optically detected magnetic resonance (ODMR). ODMR is used to find the spin resonances of the NV center, which is a preliminary step to any coherent manipulation of the NV spin, as well as a powerful sensing tool in itself.

This section also covers the experimental setup used for this operation and for many other NV experiments, as well as the concept of longitudinal and transverse magnetic field with regard to the NV principal axis.

\subsection{Experimental setup}
\label{Sec setup cahp 1}
\begin{figure}[h!]
\centering
\includegraphics[width=0.8\textwidth]{chapter 1/Figures/shema_exp}
\caption{Confocal microscopy setup (details in main text).}
\label{setup ch1}
\end{figure}
The three most common experimental degrees of freedom to consider for NV experiments are: the optical illumination, the microwave field and the external magnetic field. Other parameters such as the crystal temperature, the electric field or the crystal stress can also be controlled for specific applications.

Finally the magnetic field is generated either with a permanent magnet (up to $\sim$ 200 mT) or with a homemade electromagnet (up to $\sim$ 30 mT) alimented via a low freqency amplifier (Leybold power function generator 522663). Both the permanent magnet and the electromagnet can be  attached to a rotation stage for precise alignment.

The X-series and the pulseblaster, as well as several instruments (microwave generator, motorized stages) are piloted by python scripts which I developed during my PhD.
%Salaud de Gabriel
%\subsection{Rabi Oscillations}
%\begin{figure}[h!]
%\centering
%\includegraphics[width=\textwidth]{chapter 1/Figures/Rabi}
%\caption{Rabi oscillations. (a) pulse sequence used to measure Rabi oscillations. (b) Experimental data on sample CVD-pink (see sample details in appendix \ref{Appendix samples}).}
%\label{Rabi}
%\end{figure}
%
%The general technique for experiments involving NV spins proceeds as follow: 
%\begin{itemize}
%\item First the spins are polarized. This is done with a green laser pulse ($1-100$ $\mu$s depending on the laser intensity) which polarizes the spins in the $\ket{m_s=0}$ state at $\sim$ 80 \%.
%\item  Then the spins are manipulated. This process often involves a microwave field resonant with one of the spin transitions.
%\item Finally the spins are read out. The simplest spin readout for NV centers consists in looking at the PL right after the laser is turned on: if the spin was in the $\ket{m_s=0}$ state, the initial PL will be high, whereas if the spin was in a $\ket{m_s=\pm 1}$ state, the initial PL will be low. For longer exposure time, the spins end up re-polarized in the $\ket{m_s=0}$ state and no information can be gained on the initial spin state anymore. The optimal readout time is of the same order as the spin polarization time ($1-100$ $\mu$s).
%\end{itemize}
%
%Fig. \ref{Rabi}-a) shows the pulse sequence used to observe Rabi oscillations with NV centers. In this case, the spins manipulation simply consists in a microwave pulse of variable duration $\tau$. The microwave frequency or the magnetic field needs to be adjusted so that at least one NV spin transition (either $\ket{0} \to \ket{+1}$ or $\ket{0} \to \ket{-1}$) is resonant with the microwave field. 
%
%Fig. \ref{Rabi}-b) shows the result of such an experiment on a spin ensemble. The PL signal oscillations correspond to the oscillations of the spin population between the bright $\ket{0}$ state and the dark $\ket{-1}$ state. For long microwave pulses ($>$ 1 $\mu$s typically), the spins reach a steady state with  $\sim$ 50 \% population in both the $\ket{0}$ and $\ket{-1}$ states.

\subsection{Optically detected magnetic resonance}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/ODMR_basique}
\caption{ODMR protocol. (a) Pulse sequence used for CW ODMR: the microwave is kept on during the whole sequence while the frequency is scanned from $f_{\rm start}$ to $f_{\rm end}$. (b) ODMR spectrum with no magnetic field on sample ADM-15-4. (see sample details in appendix \ref{Appendix samples}).}
\label{ODMR basique}
\end{figure}

Whenever the microwave becomes resonant with one of the NV transition, it induces a depolarization of the NV center which results in a drop of PL. This allows an optical reading of the NV spin spectral response.

Fig. \ref{ODMR basique}-b) shows an example of ODMR spectrum on an NV ensemble with no external magnetic field. The reason as to why there are two peaks instead of a single one centered on $D=2870\ \rm MHz$ is the splitting of the $\ket{+1}$ and $\ket{-1}$ levels occasioned by the strain and local electric field, which will be further discussed in chapter 4. The asymmetry between the two peaks could come either from the microwave polarization or from the uneven spectral response of the antenna \footnote{Our antennas are not impedance matched.}.

ODMR contrast of up to 30 \% of the PL have been recorded with single spins \citep{dreau2011avoiding}. We typically observe ODMR contrast $\leq 5\ \%$ with NV ensemble.

\subsubsection{ODMR with lock-in detection}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/ODMR_lockin}
\caption{ODMR protocol with lock-in detection. (a) Simplified confocal microscope setup with a lock-in amplifier (LIA) modulating the mircowave and demodulating the PL. (b) Demodulated ODMR spectrum with no magnetic field on sample ADM-15-4 (see sample details in appendix \ref{Appendix samples}).}
\label{ODMR lockin}
\end{figure}

The ODMR protocol is well suited for lock-in detection with a modulation of the microwave amplitude or frequency modulation \footnote{These are preferred to laser modulation as only the spin part of the NV centers is affected by the microwave field}.

The principle of lock-in detection will not be covered here, the key idea is that lock-in detection allows to translate the signal spectrally in a region less prone to noise. In this case, the ODMR signal is a DC measurement and is therefore prone to slow freqency noise such as electric noise, thermal or mechanical drifts and laser amplitude fluctuations. The microwave field can thus be modulated (in amplitude, frequency or phase) at a frequency fast enough to mitigate the low frequency noise, but slow enough compared to the NV photodynamics.

\subsection{ODMR with NV ensemble: longitudinal and transverse magnetic field}
\label{sec champs transverse}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/ODMR_8_classes}
\caption{(a) Geometry of the NV center and definition of $B_z$ and $B_\perp$. (b) Representation of the 4 possibles orientations (``classes") of the N$-$V axes, corresponding to the 4 $\langle 111 \rangle$ directions of the diamond cubic cell. (c) Simulation of the evolution of the transition frequencies for each class of NV centers as a function of the amplitude of the applied magnetic field ($\textbf{B}$ orientation chosen arbitrarily). (d) ODMR spectrum on sample ADM-150-1 (see sample details in appendix \ref{Appendix samples}) for a magnetic field of $\sim 90\ \rm G$.}
\label{ODMR 8 classes}
\end{figure}

We mentioned previously that the NV spin Hamiltonian (eq. (\ref{NV spin Hamiltonian basic})) was anisotropic due to the fine structure term $D S_z^2$ where $z$ is the direction of the N$-$V axis in the crystal. The magnetic field is thus often decomposed as a sum of a longitudinal magnetic field $B_z$ and a transverse magnetic field $B_\perp=\sqrt{B_x^2+B_y^2}$ (see Fig. \ref{ODMR 8 classes}-a).

For weak magnetic fields ($\gamma |\textbf{B}| \ll D$), the energy of each of the three spin Hamiltonian eigenstates $\ket{0}$, $\ket{+1}$ and $\ket{-1}$\footnote{The notation $\ket{0}$, $\ket{\pm 1}$ to designate the spin Hamiltonian eigenstates is slightly abusive as $S_z$ does not commute with $\mathcal{H}$ when $B_\perp \neq 0$. The true eigenstates however remain close to $\ket{0}$, $\ket{\pm 1}$ as long as $\gamma_e B_z > \frac{(\gamma_e B_\perp)^2}{D}$.} can be approximated in a perturbative approach by: %cf AN.py
\begin{align}
\nu_0&= -\frac{(\gamma_e B_\perp)^2}{D}, \nonumber \\
\nu_{-1} &= D - \gamma_e B_z + \frac{(\gamma_e B_\perp)^2}{2D}, \nonumber \\
\nu_{+1} &= D + \gamma_e B_z + \frac{(\gamma_e B_\perp)^2}{2D}.
\end{align}

Fig. \ref{ODMR 8 classes}-b) represent the 4 possible orientation of the N$-$V axis in a diamond single crystal. We refer to these orientations as ``classes" since NV centers belonging to the same class will perceive the same transverse and longitudinal magnetic field, and thus will have the same spin Hamiltonian. For large NV ensembles (a confocal spot can contain more than $10^9$ NV centers in dense ensemble), we can consider that each class is equally represented \footnote{It is possible to create NV centers with a preferential orientation \citep{lesik2014perfect} but this requires specific conditions during the growth process of the diamond.} and that each class gives rise to two lines in the ODMR spectrum (one for the $\ket{0} \to \ket{-1}$ transition and one for the $\ket{0} \to \ket{+1}$ transition).

Fig. \ref{ODMR 8 classes}-c) shows a simulation of the evolution of the transition frequencies $\nu_{+1} - \nu_{0}$ and $\nu_{-1} - \nu_{0}$ for all 4 classes as a function of the magnetic field amplitude, the magnetic field orientation being held to an arbitrary direction. Fig. \ref{ODMR 8 classes}-d) shows an ODMR spectrum with NV ensembles. 8 lines are visible corresponding to the two transitions of the 4 classes. From the position of these 8 lines, it is possible to determine the projection of the magnetic field on each of the 4 N$-$V axes, and thus to reconstruct the full 3D magnetic field with respect to the diamond crystalline axes.

There are different relaxation times corresponding to different physical aspects. The longitudinal relaxation time $T_1$ corresponds to the characteristic time of the return to thermal equilibrium for the populations (diagonal elements of the density matrix). The transverse relaxation times $T_2$ and $T_2^*$ correspond to the relaxation of the coherences  (off-diagonal terms of the density matrix).

$T_2$ and $T_2^*$ are called the homogeneous and inhomogeneous dephasing times\footnote{We follow here the NMR nomenclature.}. $T_2^*$ represents to the absolute dephasing time of the spin (or spin ensemble) coherence, while $T_2$ represent the dephasing time measured when a dynamical decoupling scheme is used (Hahn echo \citep{hahn1950spin}, XY-n \citep{maudsley1986modified}, CPMG \citep{carr1954effects} , ...). While the $T_2^*$ value is absolute, the $T_2$ value depend on the precise decoupling scheme used. $T_2$ and $T_2^*$ values are ultimately limited by the spin lifetime:
\begin{equation}
T_2^* < T_2 < 2 T_1.
\end{equation}

The relaxation times of NV centers depend mainly on their environment, and can vary by several order of magnitude depending on the employed sample. The following paragraphs will present the measurement protocols of the different relaxation times and the typical values for the samples used in this manuscript.

\subsection{NV spin $T_1$}
\label{sec T1 NV}
\subsubsection{Measurement protocol}
\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/T1_basique}
\caption{Basic $T_1$ measurement scheme. (a) Pulse seqeunce used. (b) Experimental data obtained on sample CVD-pink (see sample details in appendix \ref{Appendix samples}.) for low laser intensity. The orange line corresponds to an exponential fit with time constant $\tau=4.0\ \rm ms$.}
\label{T1 basique}
\end{figure}

Fig. \ref{T1 basique} shows a basic protocol to measure the NV $T_1$ time. Fig. \ref{T1 basique}-a) shows the pulse sequence used, which in this case does not involve a microwave field, and Fig. \ref{T1 basique}-b) shows the measured PL obtained by varying the dark time $\tau$. The curve can then be fitted to obtain the spin ensemble $T_1$ value.

\begin{figure}[h!]
\centering
\includegraphics[width=\textwidth]{chapter 1/Figures/T1_protocole_soustraction}
\caption{Common mode rejection $T_1$ measurement scheme. (a) Pulse sequence. (b) PL signals $S_1$ and $S_2$ obtained on sample CVD-pink with a high laser intensity ($I_{\rm las}=5\ \rm mW$). (c) Subtraction $S_1-S_2$ versus dark time $\tau$. The experimental data is fitted with an exponential decay of time constant $\tau=4.1\ \rm ms$.}
\label{T1 soustraction}
\end{figure}

The sequence shown in Fig.\ref{T1 basique} can be improved by implementing a common mode rejection technique described in Fig. \ref{T1 soustraction}-a). This protocol works as follow: first, a basic $T_1$ sequence is recorded, which we will denote $S_1$. Then the sequence is repeated with the addition of a $\pi$-pulse on one the NV transitions right before the readout. This $\pi$-pulse inverts the population of the $\ket{0}$ and $\ket{\pm 1}$ states. The second readout is labeled $S_2$, and the final spin state readout is $S_1 -S_2$ which rejects every contribution to the PL besides the spin relaxation.

Fig. \ref{T1 soustraction}-b) shows the experimental result of the $S_1$ and $S_2$ sequences on the same sample that was used in Fig. \ref{T1 basique}, with a higher laser intensity. We can notice a stark difference between the $S_1$ curve in Fig. \ref{T1 soustraction}-b) and the curve in Fig. \ref{T1 basique}-b), despite both of them coming from the same sample and with the same measurement protocol. The reason for the difference is that the higher laser intensity used in Fig. \ref{T1 soustraction} photoionizes some of the $NV^-$ into NV$^0$ via a two-photon absorption scheme. The NV$^0$ can then capture an electron during the dark time \citep{giri2018coupled, giri2019selective} which changes the number of NV$^-$ centers probed depending on the dark time length. The goal of the subtraction scheme is to suppress this charge state recombination noise which is common to the two sequences $S_1$ and $S_2$.

Fig. \ref{T1 soustraction}-c) shows the subtraction of the $S_1$ and $S_2$ signal. Despite the complex shape of the initial $S_1$ and $S_2$ signals, $S_1-S_2$ is satisfyingly fitted by an exponential decay similar to the results found in Fig. \ref{T1 basique}.
